How Machine Learning can Advance Theory Formation

### 2023-05-23 presentation_mpib2023.Rmd
update 2023-05-23 08:13:32

---
title: "Developmental Data Science"
author: "Caspar J. Van Lissa<sup>1</sup>"
date: "<p><sup>1</sup>: Utrecht University, dept. Methodology & Statistics | Open Science Community Utrecht<br>Funded by NWO Veni grant  VI.Veni.191G.090"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
output: 
  ioslides_presentation:
    widescreen: true
    css: slide.css
bibliography      : ["manuscript/references.bib", "../rigorous_exploration/manuscript/r-references.bib", "../rigorous_exploration/manuscript/references.bib", "../rigorous_exploration/manuscript/Rigorous_exploration.bib", "../rigorous_exploration/manuscript/Veni.bib", "../rigorous_exploration/manuscript/Veni_sysrev.bib"]
---

@@

_Machine learning_ can help advance _theory formation_ in developmental psychology

* Replication crisis [@scheelWhyMostPsychological2022; @lavelleWhenCrisisBecomes2021]
* Solutions improve deductive (theory-testing) research practices
    + questionable research practices (QRPs)
    + replication
    + preregistration

Many preregistered hypotheses are not supported [@scheelExcessPositiveResults2021]

@@

* most psychological theories fall short of this definition
    + insufficiently precise
    + too flexible
* Compare to @walasekEvolutionaryModelSensitive2022

@@

```{r, out.width="60%"}
library(tidySEM)
lay <- get_layout("",     "the1", "", 
                  "pat1", "",     "hyp1",
                  "",     "obs1", "",
                  "",     "tes",  "", rows = 4)
nodes <- data.frame(
  name = c("pat1", "the1", "obs1", "tes", "hyp1"),
  label = c("Patterns", "Theory", "Observation", "Test", "Hypothesis"),
  shape = "rect")
edges <- data.frame(
  from = c("the1", "hyp1", "obs1", "pat1", "obs1"),
  to = c("hyp1", "obs1", "pat1", "the1", "tes"),
  arrow = "last"
)
g <- prepare_graph(edges = edges, layout = lay, nodes = nodes, angle = 5)
g$edges$connect_from <- c("right", "bottom", "left", "top", "bottom")
g$edges$connect_to <- c("top", "right", "bottom", "left", "top")

plot(g)
```

@@

Role of flexibility:

* Confirmatory methods poorly suited to exploratory research
    + p-values: Type I error is 5%, but false discovery rate (FDR) is 14-50% [@vidgenPValuesMisunderstoodMisused2016].
    + Model fit indices: manually specify models, no guarantee that best model is included

"unguided exploration" is effortful and inflates the risk of spurious results

@@

```{r}
knitr::include_graphics("Figure Machine learning 4.png")
```

@@

Psychological theories rarely account for complex effects

@@

* E.g.: assumption that development follows a latent growth curve (LGC).
* SEM forests [@brandmaierTheoryguidedExplorationStructural2016]
* regularized SEM [@jacobucciRegularizedStructuralEquation2016]

@@

ggplot() +
  xlim(-3, 3) +
geom_function(
  data = NULL,
  stat = "function",
  position = "identity",
  fun = ~.x^2,
  na.rm = FALSE,
  inherit.aes = TRUE,
  size = 1,
  aes(colour = "Total error")
) +
  geom_function(
    data = NULL,
    stat = "function",
    position = "identity",
    fun = ~2^.x -2,
    na.rm = FALSE,
    inherit.aes = TRUE,
    aes(colour = "Bias"),
    size = 1
  )+
  geom_function(
    data = NULL,
    stat = "function",
    position = "identity",
    fun = ~2^-.x -2,
    na.rm = FALSE,
    inherit.aes = TRUE,
    aes(colour = "Variance"),
    size = 1
  ) +
  labs(x = "Model complexity",
       y = "Optimal model complexity",
       color = "Legend") +
  scale_color_manual(values = colors) +
  geom_vline(xintercept = 0, linetype = 2) +
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

@@

* abductive formal theory construction (AFTC) framework
    + construct formal theory, generate data, mathematically compare to empirical data
    + Discrepancies: amend formal theory

* There is a paucity of good theory
* Need for exploratory research for theory formation
* Machine learning for rigorous exploration
    + automates model building
    + incorporates checks and balances for generalizable results
* Unsupervised learning can assist in phenomenon detection
* Supervised learning to identify important predictors
* Some algorithms incorporate basic theoretical elements

# Applied examples

@@

## Knowledge gap

Substantial empirical research, but:

* Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014]
* No consistent terminology [@bariolaChildAdolescentEmotion2011]
* No conceptual frameworks [@stifterEmotionRegulation2019]
* No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019]

Knowledge must be consolidated into overarching theory

@@

First step: identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]

We conducted a text mining systematic review (TMSR) [@vanlissaMappingPhenomenaRelevant2021]

  * Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
  * TMSR: Unlimited sample size, transparent, objective, reproducible

6653 papers on Addresses **emotion regulation** in population overlapping with adolescence [10-24]

<doi.org/10.1007/s40894-021-00160-7>

@@

## Baseline network

```{r figbaseline, warning=FALSE, eval = TRUE, results='asis', out.width = "90%", fig.cap="Phenomena relevant to adolescents' emotion regulation according to theory (a) and narrative reviews (b; transparent nodes indicate constructs also present in the theory)."}
knitr::include_graphics("../veni_sysrev/baseline_network.png") 
```

@@

![](../veni_sysrev/varimps_small.png)

## Co-occurrence graph

```{r networks1, eval = TRUE, results='asis', out.width = "90%"}
knitr::include_graphics("../veni_sysrev/tmnetworks_small.png")
```

@@

* Empirical research relies on theory; under-theorized phenomena may be overlooked
    + This study offers guidance for phenomena to consider as confounders/causes
* Unembedded terms indicate promising areas of future research
* Present study serves as starting point for theory development

## Next step

@@

## Solution: Machine learning

* Allows non-linear differences between trajectories
* Performs variable selection
  + Cast wide net among potentially relevant predictors
* Checks & balances ensure generalizability
* Exploratory: No hypothesis

## Method illustrated

```{r}
knitr::include_graphics("tree.png")
```

@@

@@

## Present study

@@

```{r, out.height="100%"}
vim <- readRDS("varimp_comb_21-08-2021.RData")
vim
```

@@

* Most important predictors are routinely assessed (big 5) or overt (conflict)
  + Prime candidates for early risk assessment
  + Conflict resolution behavior can be taught
    - Target for intervention if association is causal
* Emphasis on parenting (in literature and theory) might not be justified
  + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]
* Order of predictors mirrors bioecological model

@@

## Conclusions

* Proximal factors more predictive of ER development than distal predictors
* Parenting may be less relevant than previously thought
* Personality and conflict behavior are candidates for early diagnosis
* Conflict skills training might be avenue for intervention

## References

>>>>

---
title: "How Machine Learning can Advance Theory Formation"
author: "Caspar J. Van Lissa<sup>1</sup>"
date: "<sup>1</sup>: Tilburg University, dept. Methodology & Statistics<br>Funded by NWO Veni grant  VI.Veni.191G.090"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
output: 
  ioslides_presentation:
    widescreen: true
    css: slide.css
bibliography      : ["manuscript/references.bib", "presentation/all_refs.bib"]
---

@@

_Machine learning_ can help advance _theory formation_ in the social sciences

* Replication crisis [@scheelWhyMostPsychological2022; @lavelleWhenCrisisBecomes2021]
* Current solutions improve deductive research (theory-testing)
    + questionable research practices [@wichertsDegreesFreedomPlanning2016]
    + replication
    + preregistration

Many preregistered hypotheses are not supported (56%) [@scheelExcessPositiveResults2021]

@@

* Most social scientific theories are:
    + insufficiently precise
    + too flexible
* Compare to @walasekEvolutionaryModelSensitive2022

@@

```{r, out.width="40%"}
knitr::include_graphics("presentation/empiricalcycle.png")
```

@@

* Confirmatory methods poorly suited to exploratory research
    + p-values: Type I error is 5%, but false discovery rate (FDR) is 14-50% [@vidgenPValuesMisunderstoodMisused2016].
    + Model fit indices: manually specify models, no guarantee that best model is included
* Role of flexibility: there are infinite ways to explore
* "unguided exploration" is effortful and inflates the risk of spurious results

@@

```{r, out.width="100%"}
knitr::include_graphics("Figure Machine learning 4.png")
```

# Advantages of Machine Learning

@@

Social scientific theories rarely account for complex effects

@@

* E.g.: assumption that development follows a latent growth curve (LGC).
* SEM forests [@brandmaierTheoryguidedExplorationStructural2016]
* regularized SEM [@jacobucciRegularizedStructuralEquation2016], (Ernst & Peikert 2022)

@@

ggplot() +
  xlim(-3, 3) +
geom_function(
  data = NULL,
  stat = "function",
  position = "identity",
  fun = ~.x^2,
  na.rm = FALSE,
  inherit.aes = TRUE,
  linewidth = 1,
  aes(colour = "Total error")
) +
  geom_function(
    data = NULL,
    stat = "function",
    position = "identity",
    fun = ~2^.x -2,
    na.rm = FALSE,
    inherit.aes = TRUE,
    aes(colour = "Bias"),
    linewidth = 1
  )+
  geom_function(
    data = NULL,
    stat = "function",
    position = "identity",
    fun = ~2^-.x -2,
    na.rm = FALSE,
    inherit.aes = TRUE,
    aes(colour = "Variance"),
    linewidth = 1
  ) +
  labs(x = "Model complexity",
       y = "Optimal model complexity",
       color = "Legend") +
  scale_color_manual(values = colors) +
  geom_vline(xintercept = 0, linetype = 2) +
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

@@

* abductive formal theory construction (AFTC) framework
    + construct formal theory
    + Generate data
    + Mathematically compare to empirical data
    + like a posterior predictive check in Bayes
    + Discrepancies: amend formal theory

<!-- ## Summary -->

<!-- * There is a paucity of good theory -->
<!-- * Need for exploratory research for theory formation -->
<!-- * Machine learning for rigorous exploration -->
<!--     + automates model building -->
<!--     + incorporates checks and balances for generalizable results -->
<!-- * Unsupervised learning can assist in phenomenon detection -->
<!-- * Supervised learning to identify important predictors -->
<!-- * Some algorithms incorporate basic theoretical elements -->

# Example: Text Mining Systematic Review

@@

<!-- ## Knowledge gap -->

<!-- Substantial empirical research, but: -->

<!-- * Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014] -->
<!-- * No consistent terminology [@bariolaChildAdolescentEmotion2011] -->
<!-- * No conceptual frameworks [@stifterEmotionRegulation2019] -->
<!-- * No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019] -->

<!-- Knowledge must be consolidated into overarching theory -->

@@

* First step: identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]
    + Often done by systematic review
* Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
* Introducing text mining systematic review (TMSR) [@vanlissaMappingPhenomenaRelevant2021]:
    + Unlimited sample size, transparent, objective, reproducible

6653 papers on **emotion regulation** in **adolescence** [age 10-24]

<https://doi.org/10.1007/s40894-021-00160-7>

@@

## Baseline network {.smaller}

```{r figbaseline, warning=FALSE, eval = TRUE, results='asis', out.width = "100%", fig.cap="Phenomena relevant to adolescents' emotion regulation according to theory (a) and narrative reviews (b; transparent nodes indicate constructs also present in the theory)."}
knitr::include_graphics("presentation/baseline_network.png") 
```

@@

![](presentation/varimps_small.png)

## Co-occurrence graph {.smaller}

```{r networks1, eval = TRUE, results='asis', out.width = "100%"}
knitr::include_graphics("presentation/tmnetworks_small.png")
```

@@

* Empirical research relies on theory; under-theorized phenomena may be overlooked
    + TMSR offers guidance for phenomena to consider as confounders/causes
* Unembedded terms indicate promising areas of future research
* TMSR network can serve as proto-theoretical nomological network

# Example: Predictors of emotion regulation development

## Person-centered models

@@

@@

## SEM Forest

```{r, out.width="70%"}
knitr::include_graphics("tree.png")
```

@@

## Study design

@@

```{r, out.width="90%"}
knitr::include_graphics("pcomb2.png")
```

## Variable importance 2

```{r, out.width="100%"}
knitr::include_graphics("violin.png")
```

@@

* Proximal factors more predictive of ER development than distal predictors
    + Order mirrors bioecological model
* Most important predictors are routinely assessed (big 5) or overt (conflict)
  + Prime candidates for early risk assessment
  + Conflict resolution behavior can be taught
    - Target for intervention if association is causal
* Emphasis on parenting (in literature and theory) might not be justified
  + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]

@@

# Limitations and future directions

## Limitations

1. Theories are still verbal and flexible
    * Even if ML leads to new insights, it is not clear how theories should be updated
2. Ignoring causality limits interpretability of machine learning
    * Incorrect direction of causality: meaningless results
    * Conditioning on a collider: biased results

## Future direction 1: Formal theoretical object

Formal theories are expressed in mathematical terms

* unambiguous
* logically sound
* explicates assumptions
* reveals ambiguities and gaps in knowledge
* facilitates conveying ideas to others
* applying rules of formal logic implies testable hypotheses
* refutation of hypotheses dictates specific modifications

## Characteristics of formal theory

According to Meehl:

1. phenomena to be explained
2. putative causal connections between them
3. direction of the effect
4. relative importance of predictors
5. interaction effects
6. functional form of associations

## Formal theory

* Characteristics 1-2 can be described by a causal network (DAG)
* Characteristics 3-6 refer to the nature of associations within that network

Augmented causal networks (ACN, see also Crielaard et al., 2022)

* encode both causal relationships and the nature of those relationships
* imply testable hypotheses
* shared in plain text, e.g., YAML or RDF
* version controlled
* shared conform FAIR software principles,
to be used and updated through distributed collaboration

## Learning a causal network

* Using TMSR, based on what previous authors have written
* Using causal discovery, based on large panel studies
* Using ESM data to leverage precedence in time
    + Causality requires association, precedence in time, and nonspuriousness

## Future direction 2: Functional form

Previous work: Post-hoc interpretation of marginal associations

Future direction: Use a-priori interpretable machine learning

* Symbolic regression applies deep learning or genetic algorithms to learn a mathematical formula that describes the relationship between an outcome and its predictors

## References {.smaller}

Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. Infant and Child Development. https://doi.org/10.1002/icd.2370

Van Lissa, C. J. (2022). Complementing preregistered confirmatory analyses with rigorous, reproducible exploration using machine learning. Religion, Brain & Behavior. https://doi.org/10.1080/2153599X.2022.2070254

Van Lissa, C. J., Beinhauer, L., Branje, S., & Meeus, W. (2022). Using machine learning to identify early predictors of adolescent emotion regulation development. Journal of Research on Adolescence. https://doi.org/10.31234/osf.io/rbqf9

Van Lissa, C. J. (2021). Mapping Phenomena Relevant to Adolescent Emotion Regulation: A Text-Mining Systematic Review. Adolescent Research Review. https://doi.org/10.1007/s40894-021-00160-7

Van Lissa, C. J., et al., (2022). Using machine learning to identify important predictors of COVID-19 infection prevention behaviors during the early phase of the pandemic. Patterns, 3(4), 100482. https://doi.org/10.1016/j.patter.2022.100482

##

### 2023-05-23 presentation_ut2023.Rmd
update 2023-05-23 08:13:32

* abductive formal theory construction (AFTC) framework
    + construct formal theory
    + Generate data
    + Mathematically compare to empirical data
    + like a posterior predictive check in Bayes
    + Discrepancies: amend formal theory

@@

## Study design

@@

Augmented causal networks (ACN, see also Crielaard et al., 2022)

* encode both causal relationships and the nature of those relationships
* imply testable hypotheses
* shared in plain text, e.g., YAML or RDF
* version controlled
* shared conform FAIR software principles,
to be used and updated through distributed collaboration

@@

Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. Infant and Child Development. https://doi.org/10.1002/icd.2370

Van Lissa, C. J. (2022). Complementing preregistered confirmatory analyses with rigorous, reproducible exploration using machine learning. Religion, Brain & Behavior. https://doi.org/10.1080/2153599X.2022.2070254

Van Lissa, C. J., Beinhauer, L., Branje, S., & Meeus, W. (2022). Using machine learning to identify early predictors of adolescent emotion regulation development. Journal of Research on Adolescence. https://doi.org/10.31234/osf.io/rbqf9

>>>>

* abductive formal theory construction (AFTC) framework
    + construct formal theory
    + Generate data
    + Mathematically compare to empirical data (like posterior predictive check in Bayes)
    + Discrepancies: amend formal theory

@@

## Present study

@@

Augmented causal networks (ACN)

* encode both causal relationships and the nature of those relationships
* can be shared in plain text (YAML-type) data
* imply testable hypotheses
* can be shared and updated by multiple researchers

@@

Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. Infant and Child Development, n/a(n/a), e2370. https://doi.org/10.1002/icd.2370

Van Lissa, C. J. (2022). Complementing preregistered confirmatory analyses with rigorous, reproducible exploration using machine learning. Religion, Brain & Behavior, 0(0), 1–5. https://doi.org/10.1080/2153599X.2022.2070254

Van Lissa, C. J., Beinhauer, L., Branje, S., & Meeus, W. (2022). Using machine learning to identify early predictors of adolescent emotion regulation development. PsyArXiv. https://doi.org/10.31234/osf.io/rbqf9
